Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-29 paper2.qmd
section:theoretical; process:write

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study discusses two prominent issues that can compromise the reliability and validity of CJ's trait estimates. The first issue concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second involves the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. The study then addresses these challenges by adopting a more systematic and integrated approach grounded in causal inference and Bayesian inference methods.

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical updates the Thurstonian model to address these challenges. The model integrates the theoretical core principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of pairwise comparison data. @sec-discussion discusses the implications of the findings and explores avenues for future research. Finally, @sec-conclusion summarizes the study's key insights.

@@

This section discusses the two prominent issues that can undermine the reliability and validity of CJ's trait estimates. @sec-theory-issue1 examines the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. @sec-theory-issue2 focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

@@

<!-- 1. What the section will do? -->
This section addresses the challenges outlined in @sec-theory-issues by leveraging the structural approach to causal inference to extend Thurstone's model. The extension combines the core theoretical principles from @sec-thurstone_theory with several key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. Specifically, @sec-theory-theoretical_P incorporates the theoretical principles into a *conceptual-population model*. Conversely, @sec-theory-theoretical_SC integrates the assessment design features into a *sample-comparison model*. Combining these elements helps to elucidate how all actors and processes interact during CJ experiments.

@@

<!-- 2. The structural approach to causal inference -->
The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data. This approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments. Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests [@Schuessler_et_al_2023, pp. 4]. Identification analysis helps researchers determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023, pp. 4]. Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010]. *Estimators* denote the methods or functions that transform data into an estimate [@Neal_2020], while *estimates* are the numerical values approximating the estimand. These estimates are obtained through *estimation*, which is the process that integrates data with an estimator [@Everitt_et_al_2010].

<!-- 4. example of an estimand and an estimator in CJ -->
For instance, researchers might aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students to two groups, each exposed to a different teaching method. Afterward, judges would compare pairs of written texts, generating a dichotomous outcome that reflects the relative quality of each text. In this scenario, researchers can rephrase the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Researchers would then use estimates from the BTL model, or its transformations, to approximate this estimand.

<!-- 5. But the estimator does not satisfy minimum identification properties -->
However, @sec-theory-issues provides compelling evidence that the BTL model suffers from several statistical and measurement issues, including identification problems. These issues ultimately raise concerns about the consistency of the model's estimates. An estimator is *consistent* if it converges to the "true" value of an estimand as the data size approaches infinity [@Everitt_et_al_2010]. Since identification is a prerequisite for consistency, the absence of proper identification means that researchers cannot achieve consistency, even with "infinite" and error-free data. As a result, deriving meaningful insights from finite data also becomes impossible [@Schuessler_et_al_2023, pp. 5].

<!-- 5. So, how do SCMs and DAGs support identification? -->
Luckily, SCMs and DAGs support identification analysis through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020]. Second, they depict causal relationships in an interactive, non-parametric manner. This flexibility enables feasible identification strategies without defining the variables' data types, the functional forms, or their parameters [@Pearl_et_al_2016, pp. 35].

<!-- See @sec-appendix for an overview of the statistical and causal inference concepts required for the development of this section. -->

@@

<!-- 1. The SCMs -->
Before integrating the theoretical principles from @sec-thurstone_theory, it is important to define SCMs and DAGs. SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables. These function use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

@@

<!-- 3. The begining -->
@fig-CJ1 displays CJ's outcome, discriminal difference, discriminal processes and judges' biases assumed by the Thurstone's theory. Notice that in the figure, the structural

@@

>>>>

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues that can undermine the reliability and validity of CJ's trait estimates. The first issue arises from the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second stems from the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. The study then addresses these issues by extending Thurstone's model through a systematic and integrated approach based on causal and Bayesian inference methods.

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's model to address these challenges. The model integrates the theoretical core principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of pairwise comparison data. @sec-discussion discusses the implications of the findings and explores avenues for future research. Finally, @sec-conclusion summarizes the study's key insights.

@@

This section identifies and discusses the two prominent issues that can undermine the reliability and validity of CJ's trait estimates. @sec-theory-issue1 examines the the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. @sec-theory-issue2 focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

@@

<!-- 1. The structural approach to causal inference -->
The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data. The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments.

<!-- 2. How does SCMs and DAGs are used? -->
Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests [@Schuessler_et_al_2023, pp. 4]. *Identification analysis* helps researchers to determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023, pp. 4]. Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010]. *Estimators* denote the methods or functions that transform data into an estimate, while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
For instance, researchers might aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students to two groups, each receiving a different teaching method. Judges would then compare pairs of students' written texts, producing a dichotomous outcome that reflects the relative quality of each text. In this scenario, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Researchers would then use estimates from the BTL model, or its transformations, to approximate this estimand.

<!-- 4. But why identification is so important? -->
However, @sec-theory-issues presents strong evidence that the BTL model has several statistical and measurement limitations. These limitations hinder the model's ability to identify distinct estimands relevant to CJ inquiries, including the one presented in the previous paragraph. Identification is crucial because it is a necessary condition for ensuring consistent estimators. Without identification, researchers cannot achieve consistency even with "infinite" and error-free data. In other words, the estimates derived from the estimator fail to converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010]. As a result, deriving meaningful insights from finite data also becomes unattainable [@Schuessler_et_al_2023, pp. 5].

<!-- 5. So, how do SCMs and DAGs support identification? -->
Luckily, SCMs and DAGs support identification analysis through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020]. Second, they depict causal relationships in an interactive, non-parametric way. This flexibility enables feasible identification strategies without requiring specification of the types of variables, the functional forms relating them, or the parameters of those functional forms [@Pearl_et_al_2016, pp. 35].

<!-- 5. What the section will do? -->
As a result, this section addresses the identified limitations by extending Thurstone's model through a systematic and integrated approach based on the structural approach to causal inference. Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments. These features include the selection of judges, stimuli, and comparisons. Combining these components clarifies the interactions among all actors and processes involved in CJ experiments. Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into a *conceptual-population model*. Meanwhile, @sec-theory-theoretical_SC integrates the assessment design features into a *sample-comparison model*.

@@

@@

<!-- 1. The SCMs -->
Before integrating the theoretical principles from @sec-thurstone_theory, it is important to define SCMs and DAGs. SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables.

For instance, @fig-CJ1_scm display the SCM relating the CJ outcome $(O_{kjabi_{1}i_{2}})$, discriminal difference $(D_{kjabi_{1}i_{2}})$, perceived discriminal processes for each text within each student $(PT_{kjai_{1}}, PT_{kjbi_{2}})$, the "true" discriminal processes for each text within each student $(T_{kjai_{1}}, T_{kjbi_{2}})$, and the judges' biases $(B_{kj})$. Notice that the structural equations use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

@@

@@

<!-- These estimates are obtained through *estimation*, which is the process that integrates data with an estimator [@Everitt_et_al_2010]. -->
